Extending statistical models for source attribution of zoonotic diseases: A study of campylobacteriosis

### 2018-03-18 paper/finalish/article_1023.tex
update tables to use consistent STs and commentary in results

\begin{table}
  \begin{center}
    \begin{tabular}{crrrrrrr}
      \toprule
      ST & \textit{asp} & \textit{gln} & \textit{glt} & \textit{gly} & \textit{pgm}& \textit{tkt}& \textit{unc}\\ \midrule
      45  & 4 & 7 & 10 & 4 & 1 & 7 & 1\\
      474 & 2 & 4 & 1 & 2 & 2 & 1 & 5 \\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The allelic profile of genotypes ST-45 and ST-474 is composed of seven allele numbers at each of the seven loci.}
  \label{tab0}
\end{table}

@@

For the pathogen \textit{Campylobacter} which causes campylobacteriosis, a worldwide gastrointestinal disease in humans, the commonly used 7-gene MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit). An illustrative example of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab0}). It shows that the genotypes ST-45 and ST-474 have different allelic combinations across the seven loci. The different allelic profiles enable comparison of gene similarities or dissimilarities so that an association between sources and infected cases can be made, by comparing the distribution of genotypes from human cases with those from potential reservoirs.

@@

The source attribution study described here uses MLST data collected in the Manawatu region of New Zealand from March, 2005 to December, 2014. The sampling was carried out over the same time period and from the same geographical location, providing $H+N=3,480$ isolates taken from human and non-human cases, $H=1,400$ of which are from humans. The non-human isolates are sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, we therefore put them into four groups representing major sources of infection: poultry, ruminants, water and others. The total number of unique STs typed from all isolates is $I=355$, and 35\% of genotypes are found among human cases. Table~(\ref{tab1}) lists five common genotypes found in human and source isolates. The first four genotypes are frequently observed in human cases. For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-61 are typically found in ruminants. Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-61 are more likely attributable to ruminants \cite{Muell, Coll, Cart}. Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or wild birds.

@@

\begin{table}
  \begin{center}
    \begin{tabular}{rrrrrr}
      \toprule
      ST & Human & Poultry & Ruminants & Water & Others\\ \midrule
      42 & 54 & 7 & 52 & 10 & 2\\
      45  & 141 & 154 & 10 & 21 & 54\\
      61 & 64 & 7 & 54 & 2 & 1\\
      474 & 239 & 60 & 15 & 5 & 7\\
      2381 & 0 & 0 & 0 & 21 & 3\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The frequency of five genotypes found from human and four source isolates.}
  \label{tab1}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{Figures/attr_all.pdf}
\caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas. The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig2}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.9\linewidth]{Figures/attr_2009.pdf}
\caption{Posterior mean attribution ($F$) of human cases from the post-intervention time period (2008-2014) with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area. The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig3}
\end{figure}

To investigate the association between source attribution and human infection, we apply the methods demonstrated in the preceding sections to the campylobacteriosis data from the Manawatu in New Zealand. The posterior mean of samples $F$ with the 80\% highest posterior density (HPD) interval for each source is illustrated on each rurality grade in Figure~(\ref{fig2}). Given different genotype distributions, the upper panel of the figure is based on the linear model, whereas the lower panel is generated from the categorical model.

For the linear model, the simple model estimates 75\% and 32\% of cases dwelling in main urban areas have acquired the infection from poultry and ruminants, respectively, whereas the asymmetric Island model provides less than 70\% and approximately 30\% of cases attributable to the respective source. Further, the 80\% HPD interval of $F$ for poultry and for ruminants based on the simple model seems marginally wider than that for the asymmetric Island model. So, there is no significant difference in the proportion of human cases attributed to each source between the asymmetric Island and the simple models. For the categorical model, the comparison between the asymmetric Island and the simple models is quite similar to that for the linear model except for the distinguishable attribution for poultry and ruminants in highly rural areas and the much wider credible intervals of $F$ for four sources. For the two fitted models, the narrower credible intervals noticed in very rural areas from the asymmetric Island model might indicate that the asymmetric Island model gives a little more accurate sampling distribution for poultry and ruminants genotypes especially typed in humans living in rural areas. Further, the relatively larger credible intervals observed in the categorical model might be the effect of allocating the number of cases to each qualitative scale (as a dummy variable), particularly in rural areas due to small observations.

Figure~(\ref{fig2}) shows that poultry starts becoming the dominant source contributable to the infection between rural areas with moderate and with high urban influence. In contrast, ruminants are becoming the most attributable source when locations are getting farther away from rural areas with a moderate urban influence. As to water and other sources, they seem to be less influenced by the rurality levels because the probabilities remain less than 0.2 over the seven categories.

Moreover, with a view to examining the robustness of our results, the approach is also applied to data with the time period of post-intervention from 2008 to 2014. The source attribution at different rurality levels is then illustrated in Figure~(\ref{fig3}), displaying that the overall attribution over four sources is moderately consistent with the patterns we previously discovered. However, Figure~(\ref{fig3}) shows a marked decline of attribution probability for poultry and a small rise in probability for ruminants in main urban areas. Despite the fact that the intervention did not eliminate the infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.

>>>>

\begin{table}
  \begin{center}
    \begin{tabular}{crrrrrrr}
      \toprule
      ST & \textit{asp}A & \textit{gln}A & \textit{glt}A & \textit{gly}A & \textit{pgm}& \textit{tkt}& \textit{unc}A\\ \midrule
      403 & 10 & 27 & 16 & 19 & 10 & 5 & 7\\
      2026 & 10 & 1 & 16 & 19 & 10 & 5 & 7\\
      474 & 2 & 4 & 1 & 2 & 2 & 1 & 5 \\
      2343 & 2 & 4 & 5 & 2 & 10 & 1 & 5\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The allelic profile of genotypes ST-45 and ST-474 is composed of seven allele numbers at each of the seven loci.}
  \label{tab:mlst}
\end{table}

@@

For the pathogen \textit{Campylobacter} which causes campylobacteriosis, a worldwide gastrointestinal disease in humans, the commonly used 7-gene MLST scheme consists of housekeeping genes \textit{asp}A (aspartase A), \textit{gln}A (glutamine synthetase), \textit{glt}A (citrate synthase), \textit{gly}A (serine hydroxymethyltransferase), \textit{pgm} (phosphoglucomutase), \textit{tkt} (transketolase), and \textit{unc}A (ATP synthase $\alpha$ subunit). An illustrative example of MLST data for \emph{Campylobacter jejuni} is presented in Table~(\ref{tab:mlst}). It shows that the genotypes ST-2026 and ST-474 have different allelic combinations across all seven loci, while ST-403 differs from ST-2026 only at \textit{gln}A, and ST-2343 differs from ST-474 at \textit{glt}A and \textit{pgm}. The different allelic profiles enable comparison of gene similarities or dissimilarities so that an association between sources and infected cases can be made, by comparing the distribution of genotypes from human cases with those from potential reservoirs.

@@

The source attribution study described here uses MLST data collected in the Manawatu region of New Zealand from March, 2005 to December, 2014. The sampling was carried out over the same time period and from the same geographical location, providing $H+N=3,480$ isolates taken from human and non-human cases, $H=1,400$ of which are from humans. The non-human isolates are sampled from chicken carcasses, cattle, sheep, environmental water, wild birds, and so on, we therefore put them into four groups representing major sources of infection: poultry, ruminants, water and others. The total number of unique STs typed from all isolates is $I=355$, and 35\% of genotypes are found among human cases. Table~\ref{tab:counts} lists five common genotypes found in human and source isolates. The first four genotypes are frequently observed in human cases. For four sources, ST-45 and ST-474 are commonly detected in poultry, while ST-42 and ST-2026 are typically found in ruminants. Some studies also discover that human cases with ST-45 or ST-474 are likely attributable to poultry, whereas human cases with ST-42 or ST-2026 are more likely attributable to ruminants \cite{Muell, Coll, Cart}. Furthermore, it also shows that there are no observations of ST-2381 in humans, poultry or ruminants, indicating that the genotype may only appear in the water or wild birds.

@@

\begin{table}
  \begin{center}
    \begin{tabular}{rrrrrr}
      \toprule
      ST & Human & Poultry & Ruminants & Water & Others\\ \midrule
      42 & 54 & 7 & 52 & 10 & 2\\
      45  & 141 & 154 & 10 & 21 & 54\\
      474 & 239 & 60 & 15 & 5 & 7\\
      2026 & 37 & 0 & 46 & 2 & 5\\
      2381 & 0 & 0 & 0 & 21 & 3\\
      \bottomrule
    \end{tabular}
  \end{center}
  \caption{The frequency of five genotypes found from human and four source isolates.}
  \label{tab:counts}
\end{table}

@@

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures/attr_all.pdf}
\caption{Posterior mean attribution ($F$) of human cases with 80\% credible intervals for source: poultry (red), ruminants (blue), water (green) and others (purple) over the rurality scales from highly rural area to main urban areas. The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig2}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures/genotype_figure.pdf}
\caption{Posterior probability for each source for four sequence types from the asymmetric Island and Dirichlet models, assuming that each source is \emph{apriori} equally likely.}
\label{fig:genotype_dist}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures/attr_2009.pdf}
\caption{Posterior mean attribution ($F$) of human cases from the post-intervention time period (2008-2014) with 80\% credible intervals for the four sources over the rurality scales from highly rural area to main urban area. The attribution is generated from both the linear and the categorical models, given the sampling distribution of genotypes with evolutionary information (the asymmetric Island model) or without any genetic information (the Dirichlet model).}
\label{fig3}
\end{figure}

To investigate the association between source attribution and human infection, we apply the methods described in the preceding sections to the campylobacteriosis data from the Manawatu in New Zealand. The posterior mean of $F$ with 80\% credible intervals for each source are illustrated for each rurality grade in Figure~\ref{fig2}. The graphs are categorized by the type of genetic model (asymmetric Island or Dirichlet) and the manner in which the rurality variable is modelled (categorical or linear on the logit scale).

Overall, the attribution results are relatively stable irrespective of the genetic model or how rurality is represented in the attribution model. The majority of human cases are attributed to ruminants and poultry, with more cases attributed to ruminants in rural areas and more cases attributed to poultry in urban areas. For both models of the genotype distribution, the linear and categorical models of rurality show broadly the same trend, suggesting that the additional flexibility given by the categorical model is not required, and that the shift in attribution as rurality changes is adequately modelled by a linear trend on the logit scale. The linear model has the advantage of tighter credible intervals as it can share data across the 7 levels of rurality, resulting in a clearer separation of ruminant and poultry attribution, particularly in highly rural areas where the data are sparse.

There are some small differences between the genotype models, with the Dirichlet model showing a greater attribution to poultry (ranging from 40\% in highly rural areas to 70\% in main urban centres) than the asymmetric Island model (ranging from 30\% in rural areas to 65\% in urban centres). This occurs in both the linear and categorical models. Interestingly, the asymmetric Island model attributes approximately 7\% of human cases across all rurality levels to sources other than poultry, ruminants and water, and gives a small attribution to water in highly rural areas, whilst the Dirichlet model indicates that both these sources are unimportant.

The small differences in attribution observed between the asymmetric Island and Dirichlet models may be due to the additional genetic information available to the asymmetric Island model. To illustrate this, Figure~\ref{fig:genotype_dist} shows the probability of source given genotype for a selection of four genotypes, assuming \emph{apriori} that that each source was equally likely. Genotypes ST-403 and ST-2343 are observed primarily in humans ($n=7$ and $n=6$ respectively) while only being observed once each among the source isolates, so that the Dirichlet model has little information available to distinguish between sources. The asymmetric Island model, however, can exploit the genetic relationship between genotypes. ST-403 differs at just one locus from ST-2026, a type observed frequently in human cases and ruminant isolates, while ST-2343 differs at two loci from common genotype ST-474, observed frequently in human cases and poultry isolates (Tables \ref{tab:mlst} and \ref{tab:counts}). Thus, the asymmetric Island model can clearly assign ST-403 to ruminants and ST-2343 to poultry, whilst the Dirichlet model cannot distinguish between sources.

As noted previously, a major public health initiative in 2007 led to a significant reduction in the number of cases of campylobacteriosis in New Zealand. In order to examine the effects of this change on attribution probabilities, we repeated the analysis using only data from 2008 to 2014. Figure~\ref{fig3} shows that the general trend in attribution by rurality is consistent with that seen with all the data, with ruminant associated campylobacteriosis in highly rural areas to poultry associated cases in urban centers linked by a linear trend. However, the proportion attributed to poultry across all rurality levels decreased (and thus attribution to ruminants increased) in this post-intervention dataset, being most marked in urban centers. Thus, although the intervention did not eliminate infection arising from poultry \cite{MuelM}, the reduction highlights the significant improvement in contribution of poultry to the disease, particularly in urban areas where most cases lived.
